Unlocking the Potential of microRNAs: Machine Learning Identifies Key Biomarkers for Myocardial Infarction Diagnosis

### 2023-08-07 Complete-V1/Complete-V1.tex
The version after revision

\title{Machine Learning Deriven Set of microRNAs as a Novel Biomarker
for Myocardial Infarction Diagnosis }

\author{  Mehrdad Samadishadlou \and  Reza Rahbarghazi \and  Farhad
Bani \and  }

@@

\institute{
        Mehrdad Samadishadlou \at
     Medical Nanotechnology gruop, Tabriz University of Medical
Sciences, Tabriz, Iran \\
     \email{\href{mailto:samadishadlou@tbzmed.ac.ir}{\nolinkurl{samadishadlou@tbzmed.ac.ir}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Reza Rahbarghazi \at
     Medical Nanotechnology gruop, Tabriz University of Medical
Sciences, Tabriz, Iran \\
     \email{\href{mailto:rahbarghazir@tbzmed.ac.ir}{\nolinkurl{rahbarghazir@tbzmed.ac.ir}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Farhad Bani \at
     Medical Nanotechnology gruop, Tabriz University of Medical
Sciences, Tabriz, Iran \\
     \email{\href{mailto:bainf@tbzmed.ac.ir}{\nolinkurl{bainf@tbzmed.ac.ir}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

\begin{abstract}
MicroRNAs (miRNAs) play a crucial role in regulating adaptive and
maladaptive responses in cardiovascular diseases, making them attractive
targets for potential therapeutics. However, their potential as novel
biomarkers for diagnosing cardiovascular diseases requires systematic
evaluation. In this study, we aimed to identify a key set of miRNA
biomarkers using integrated bioinformatics and machine learning
analysis. We combined and analyzed three gene expression datasets from
the Gene Expression Omnibus (GEO) database, which contains peripheral
blood mononuclear cell samples from individuals with myocardial
infarction (MI), stable coronary artery disease (CAD), and healthy
individuals. Additionally, we selected a set of miRNAs based on their
area under the curve of the receiver operating characteristic curve
(AUC-ROC) to separate the CAD/MI samples. We designed a two-layer
architecture for sample classification, with the first layer isolating
healthy samples from not-healthy ones, and the second layer classifying
stable CAD and MI samples. We trained different machine learning models
using both biomarker sets and evaluated their performance on a test set.
We identified miR-21, miR-186, and miR-32 as the only miRNAs among the
differentially expressed genes, and a set including miR-186, miR-21,
miR-197, miR-29A, and miR-296 as the optimum set of miRNAs selected by
their AUC-ROC. Both biomarker sets could distinguish healthy from
not-healthy samples with complete accuracy. The best performance for the
classification of CAD and MI was achieved with an SVM model trained
using the biomarker set selected by AUC-ROC, with an AUC-ROC of 0.95,
and an accuracy of 0.88. Our study demonstrated that miRNA signatures
derived from peripheral blood could serve as valuable novel biomarkers
for cardiovascular diseases.
\\
\keywords{
        microRNA \and
        Machine Learning \and
        Myocardial Infarction \and
    }

@@

At present, cardiovascular diseases (CVDs) are the leading cause of
human mortality with 32\% of all global deaths. It is estimated that
about 85\% of CVDs mortality were diagnosed with myocardial infarction
(MI) ({``Cardiovascular Diseases ({CVDs})''} n.d.). MI is an acute
coronary syndrome with sudden blockage and stenosis of the coronary
artery, and subsequent myocardial ischemia, leading to extensive
cardiomyocyte damage and necrosis (Yap et al. 2023).

Over the last 50 years, numerous attempts have been collected to use
biomarkers to facilitate diagnosis, assess risk, follow-up therapy, and
determine therapeutic efficacy in CVDs candidates. Based on the released
guidelines, cardiac troponins (cTns) are used as a highly-sensitive and
accurate approach for the detection of MI. Despite the inherent
advantages, the high-rate sensitivity of cTn-based assays has also led
to more false positive results (Thygesen et al. 2018), which do
necessitate the advent and development of new modalities with
pathological values. To improve diagnostic value upon existing MI
biomarkers, the combination of complementary biological markers, such as
microRNAs (miRNAs) and other genetic factors, is proposed. Previous
researches support the notion that miRNAs exhibit the great potential to
be used as alternative biomarkers in CVDs detection and follow-up
(Schulte et al. 2020). It is suggested that miRNAs possess 18-22
nucleotides and can play a crucial role in the regulation of gene
expression. Evidence point to the fact that miRNAs are involved in the
pathogenesis of cardiac tissue injury and could be considered as
theranostics targets in terms of CVDs. These elements can easily
circulate in biofluids with both diagnostic as well as prognostic values
(Schulte, Karakas, and Zeller 2017). Several biological activities such
as angiogenesis, cardiomyocyte growth and contractility, lipid
metabolism, plaque formation, and cardiac rhythm are regulated by miRNAs
(Kalayinia et al. 2021). It is postulated that the function and
diagnostic properties of miRNAs are beyond the myocardium in CVD
patients. To be specific, the expression of miRNAs can vary in different
biofluids and cell components such as serum and peripheral blood
mononuclear cells (PBMCs) (Soler-Botija, Gálvez-Montón, and Bayés-Genís
2019).

PBMCs are a fraction of white blood cells (WBCs), including monocytes,
lymphocytes, macrophages, and other cells belonging to the immune system
(Gao et al. 2020). Emerging data have indicated that PBMCs can be used
as a valid source of biomarkers for monitoring various pathological
conditions. Of note, the alteration of mRNAs and miRNAs under
pathological conditions gives us valuable information about different
kinds of disorders. PBMCs could recapitulate the conditions of the
target tissues, thus, providing a highly sensitive and specific source
of biomarkers (Mosallaei et al. 2022). Commensurate with these
conditions, these cells are repositories of dysregulated genes and
miRNAs expression profiles in CVDs (Gao et al. 2020; Mosallaei et al.
2022).

In recent years, the advent and use of machine learning (ML) is an
exciting prospect for advancing scientific discoveries. Although the
concept of ML and its initial algorithms were conceived many years ago,
recent improvements in computing power and access to vast amounts of
data have shown that ML techniques outperform classical statistical
methods in various fields. Furthermore, the progress made in omics
technologies has enabled the analysis of massive and intricate
biological data sets, consisting of hundreds to thousands of samples,
which makes it possible for ML to extract valuable biological
information from such data (Torun et al. 2023). Therefore, ML offers
novel techniques to integrate and analyze the various omics data
enabling the discovery of de novo biomarkers. These biomarkers help us
in accurate disease prediction, patient stratification, and finding new
therapeutics (Reel et al. 2021).

In this study, we aimed to identify potential miRNA biomarkers for MI
patients by combining and analyzing three different microarray datasets
from PBMCs. It is suggested that the integration of omics data with
bioinformatics and ML techniques could be a promising tool in the
discovery of new and more accurate biomarkers for monitoring MI.
Besides, this approach can deepen our vision into the underlying
mechanisms of MI and aid in the development of valid diagnostic
biomarkers, and patient stratification.

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO)
database (\url{https://www.ncbi.nlm.nih.gov/geo/}). To obtain sufficient
classification power between MI, healthy and stable coronary artery
disease (CAD) samples, a relatively large sample size was required.
Therefore, GSE59867 for MI and CAD samples, and GSE56609 and GSE54475
for healthy samples were selected. All samples were produced using
Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. Only healthy,
stable CAD and early-stage MI samples were selected from these datasets
for further analyses. The basic information for the three datasets
evaluated in the current study is provided in Table @ref(tab:datasets).

\begin{longtable}[]{@{}cccccc@{}}
\caption{Basic information of the GEO microarray
datasets.}\tabularnewline
\toprule()
Dataset & Platform & Healthy & CAD & MI & Refrence \\
\midrule()
\endfirsthead
\toprule()
Dataset & Platform & Healthy & CAD & MI & Refrence \\
\midrule()
\endhead
GSE59867 & GPL6244 & - & 46 & 111 & (Maciejak et al. 2015) \\
GSE56609 & GPL6244 & 46 & - & - & (Matone et al. 2015) \\
GSE54475 & GPL6244 & 5 & - & - & (Canali et al. 2014) \\
\bottomrule()
\end{longtable}

\hypertarget{pre-processing}{%
\subsection{Pre-processing}\label{pre-processing}}

Raw data (CEL files) of all datasets were downloaded from the GEO and
pre-processed using the fRMA package (M. N. McCall, Bolstad, and
Irizarry 2010). fRMA allowed to pre-process of individual microarray
samples and combining them consistently for analysis. For each dataset,
background correction was performed using the RMA algorithm and then it
was quantile normalized based on the reference distribution. During
summarization, batch effects were removed and variances of the gene
expressions were estimated by taking into account these probe-specific
effects. For those multiple probe sets matched to the identical gene,
the mean log fold change was retained. Therefore, fRMA can be seen as a
batch effect removal technique for different datasets that are produced
by identical microarray platforms. Thus, to ensure batch effect removal,
the principal component analysis and the relative log expression of
train samples were plotted before and after fRMA (Lazar et al. 2013).

@@

The barcode algorithm proposed by McCall et al. (Matthew N. McCall et
al. 2011) transformed the actual expression values into binary barcode
values. Huge sets of samples were collected and normalized using fRMA
for several platforms as well as for Affymetrix Human Gene 1.0 ST Array
(GPL6244) platform. The distribution of the expressed and unexpressed
observed intensities for each gene is estimated using these normalized
sets. Genes were considered expressed (and their value coded to 1) or
unexpressed (and their value coded to 0) according to the following
equation:

@@

where \(x_{ij}\) is the normalized intensity of gene \(i\) in sample
\(j\), \(C\) is a user-defined parameter, \(\sigma^{ne}\) is the
standard deviation of the non-expressed distribution, and \(\mu^{ne}\)
is the mean of the non-expressed distribution. The barcode
representation of a sample is a vector of ones and zeros denoting which
genes are estimated to be expressed (ones) and unexpressed (zeros). The
barcode algorithm was implemented by the barcode function in the R fRMA
package, and the default value of \(C\) was used.

To determine if the expressed ratios differed in the MI group versus the
healthy control group, Fisher's exact test for individual genes was
carried out upon the barcode values. Genes with a false discovery rate
(FDR) of \(< 0.05\), which was calculated through the Benjamini-Hochberg
(BH) procedure to adjust for multiple testing issues, were considered as
differentially expressed genes. The same procedures were conducted on
CAD versus healthy controls as well as MI versus CAD group to find the
DEGs between them.

@@

Using the R clusterProfiler package, the Kyoto Encyclopedia of Genes and
Genomes (KEGG) pathway enrichment analysis and Gene Ontology (GO)
functional annotation were carried out on the differentially expressed
genes. The GO analysis included biological process (BP), cellular
component (CC) and molecular function (MF) categories. An adjusted
p-value of less than 0.05 was considered to indicate a statistically
significant difference. Enrichments were conducted on the MI-healthy and
CAD-healthy DEGs. In these analyses, all default parameters were used.

@@

The ML analysis was performed using Python software, ver. 3.9, Numpy
(Harris et al. 2020), pandas (McKinney 2010), and Scikit-Learn packages
(Pedregosa et al. 2011). Whenever hyper-tuning was needed, the
scikit-opt package (Head et al. 2021) was used. In all ML analyses, the
datasets were divided into train and test sets by a 0.7:0.3 ratio and
all reported results are the average of 10-fold cross-validation.

Two different approaches were used for selecting miRNAs for model
training. The first approach was using the miRNAs that are
differentially expressed. In the second approach, miRNAs with individual
area under the curve of the receiver operating characteristic (AUC-ROC)
over 0.8 for separating MI from CAD were selected. Having the result of
these two different approaches can provide an informative comparison
between the predictive capabilities of sets of miRNAs selected with
different logics.

@@

In this approach, a two-layer architecture was deployed to the data to
maximize the prediction values. The first layer predicted whether a
sample is healthy or not, and the second layer separated MI from CAD in
the samples which were predicted as not healthy in the first layer. To
this end, a distinct ML model was trained for each layer. Since there is
a limited number of miRNAs in DEGs, both layers were trained with all of
them. For further comparison with the models' performance, the ROC curve
of each miRNA for classifying healthy and not-healthy, as well as CAD
and MI, were generated using a Logistic Regression model.

\hypertarget{first-layer-for-isolation-of-healthy-and-not-healthy-samples}{%
\paragraph{First layer for isolation of healthy and not-healthy
samples:}\label{first-layer-for-isolation-of-healthy-and-not-healthy-samples}}

An SVM model using RBF kernels was trained and hyper-tuned using all
miRNAs in DEGs. To handle the severe imbalance in the number of samples
(51 for the healthy and 157 for the not-healthy groups), the sample
weight for the healthy and the not-healthy samples were set to 1 and
0.5, respectively.

\hypertarget{second-layer-for-seperating-mi-and-cad-samples}{%
\paragraph{Second layer for seperating MI and CAD
samples:}\label{second-layer-for-seperating-mi-and-cad-samples}}

For the sake of reaching the highest classification performance using
the set of miRNAs, different models were investigated. To do so, SVM
(with linear, polynomial, and RBF kernels), Logistic Regression (LR),
Random Forests (RF), k-Nearest Neighbor (kNN), Gradient Boosting (GB),
XGBoost (XGB) and Decision Tree (DT) models were trained. All models
were trained with their pre-set parameters with 10-fold
cross-validation. The criteria for choosing the best model were the
highest accuracy and AUC-ROC on the test set. The best model was
hyper-tuned with the scikit-opt package (Head et al. 2021) to get the
best predictive performance.

@@

Like the previous approach, a two-layers strategy was conducted. The
first layer classified samples into healthy and not-healthy, and the
separated MI and CAD samples. However, to keep the number of miRNAs as
low as possible miRNAs were selected from the second layer, and then
their performance wase evaluated in the first layer. AUC-ROC of all
miRNAs for classifying MI and CAD samples were calculated and the miRNAs
with the AUC-ROC over 0.8 were selected. The ROC curves for each
selected miRNA for separating healthy samples from not-healthy and MI
from CAD samples were also plotted for further comparison.

\hypertarget{first-layer-for-the-detection-of-healthy-and-not-healthy-samples}{%
\paragraph{First layer for the detection of healthy and not-healthy
samples:}\label{first-layer-for-the-detection-of-healthy-and-not-healthy-samples}}

An SVM model with an RBF kernel was trained using the selected set of
miRNAs. Additionally, the model was hyper-tuned to find the
hyper-parameters for the highest AUC-ROC and accuracy. The ROC curve and
confusion matrix for the best model were reported.

\hypertarget{second-layer-for-separating-mi-and-cad}{%
\paragraph{Second layer for separating MI and
CAD:}\label{second-layer-for-separating-mi-and-cad}}

The selected miRNAs set was used to train different algorithms to find
the best model. Similar to the previous approach, SVM (with linear,
polynomial, and RBF kernels), LR, RF, kNN, GB, XGB, and DT were trained.
All models were trained with their pre-set parameters using 10-fold
cross-validation. The models with the highest AUC-ROC and accuracy on
the test set were selected and hyper-tuned using the scikit-opt package
(Head et al. 2021). The ROC curve and confusion matrix for the best
model were reported.

@@

\hypertarget{pre-processing-1}{%
\subsection{Pre-processing}\label{pre-processing-1}}

The PCA plots of the training samples are shown in Figure @ref(fig:PCA)A
and B. As shown, healthy samples were separated from CAD or MI samples
in primary data and also after conducting fRMA. In the RLE plot, there
was a distinct difference between dataset means for all samples before
conducting fRMA (Figure @ref(fig:PCA)C). All datasets were rearranged
around 0 in the RLE plot after conducting fRMA (Figure @ref(fig:PCA)D).
Moreover, there was a clear change in inter-quantile distances, but the
values still were over 0.1.

\begin{figure}
\includegraphics[width=1\linewidth]{../../DEG and DEM/Pics/PCA and RLE of trains} \caption{PCA and RLE plot for all samples before and after fRMA.}\label{fig:PCA}
\end{figure}

@@

According to the cutoff criterion of \(FDR < 0.05\), there were 860 DEGs
between the MI and the healthy samples. Among them, 323 were
up-regulated, and 537 were down-regulated in MI compared to the healthy
controls. In CAD and healthy groups comparison, we found 670 DEGs, of
which 262 and and 408 DEGs were up- and down-regulated, respectively in
CAD samples. In the MI and CAD groups, the number of DEGs was 260, and
the number of up- and down-regulated genes in MI samples were 144 and
116, respectively in comparison with CAD samples. These data are
summarized in Table @ref(tab:DEGstab).

@@

{\centering \includegraphics[width=0.45\linewidth]{../../DEG and DEM/Pics/venny}

@@

\caption{Venn diagram for DEGs in CAD/Healthy, MI/Healthy, and MI/CAD comparison.}\label{fig:venn}
\end{figure}

The Venn diagram in Figure @ref(fig:venn) shows that CAD and MI samples
share a majority of their DEGs. From 860 DEGs of MI/healthy and 670 DEGs
of CAD/healthy, 531 genes were common which is 62\% of MI/healthy DEGs
and 79\% of CAD/healthy DEGs.

>>>>

\title{Unlocking the Potential of microRNAs: Machine Learning Identifies
Key Biomarkers for Myocardial Infarction Diagnosis }

\author{  Mehrdad Samadishadlou \and  Reza Rahbarghazi \and  Zeynab
Piryaei \and  Mahdad Esmaeili \and  Çığır Biray Avcı \and  Farhad
Bani \and  Kaveh Kavousi \and  }

@@

\institute{
        Mehrdad Samadishadlou \at
     Medical Nanotechnology gruop, Tabriz University of Medical
Sciences, Tabriz, Iran \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Reza Rahbarghazi \at
     Medical Nanotechnology gruop, Tabriz University of Medical
Sciences, Tabriz, Iran \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Zeynab Piryaei \at
     Laboratory of Complex Biological Systems and Bioinformatics (CBB),
Department of Bioinformatics, Institute of Biochemistry and Biophysics
(IBB), University of Tehran, Iran. \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Mahdad Esmaeili \at
     Medical Bioengineering Department, School of Advanced Medical
Sciences, Tabriz University of Medical Sciences, Tabriz, Iran \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Çığır Biray Avcı \at
     Medical Biology Department, School of Medicine, Ege University,
İzmir, Turkey \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Farhad Bani \at
     Medical Nanotechnology gruop, Tabriz University of Medical
Sciences, Tabriz, Iran \\
     \email{\href{mailto:bainf@tbzmed.ac.ir}{\nolinkurl{bainf@tbzmed.ac.ir}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Kaveh Kavousi \at
     Laboratory of Complex Biological Systems and Bioinformatics (CBB),
Department of Bioinformatics, Institute of Biochemistry and Biophysics
(IBB), University of Tehran, Iran \\
     \email{\href{mailto:kkavousi@ut.ac.ir}{\nolinkurl{kkavousi@ut.ac.ir}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

\begin{abstract}
MicroRNAs (miRNAs) play a crucial role in regulating adaptive and
maladaptive responses in cardiovascular diseases, making them attractive
targets for potential biomarkers. However, their potential as novel
biomarkers for diagnosing cardiovascular diseases requires systematic
evaluation. In this study, we aimed to identify a key set of miRNA
biomarkers using integrated bioinformatics and machine learning
analysis. We combined and analyzed three gene expression datasets from
the Gene Expression Omnibus (GEO) database, which contains peripheral
blood mononuclear cell (PBMC) samples from individuals with myocardial
infarction (MI), stable coronary artery disease (CAD), and healthy
individuals. Additionally, we selected a set of miRNAs based on their
area under the receiver operating characteristic curve (AUC-ROC) for
separating the CAD and MI samples. We designed a two-layer architecture
for sample classification, in which the first layer isolates healthy
samples from unhealthy samples, and the second layer classifies stable
CAD and MI samples. We trained different machine learning models using
both biomarker sets and evaluated their performance on a test set. We
identified miR-21, miR-186, and miR-32 as the only miRNAs among the
differentially expressed genes, and a set including miR-186, miR-21,
miR-197, miR-29A, and miR-296 as the optimum set of miRNAs selected by
their AUC-ROC. Both biomarker sets could distinguish healthy from
not-healthy samples with complete accuracy. The best performance for the
classification of CAD and MI was achieved with an SVM model trained
using the biomarker set selected by AUC-ROC, with an AUC-ROC of 0.96 and
an accuracy of 0.94 on the test data. Our study demonstrated that miRNA
signatures derived from PBMCs could serve as valuable novel biomarkers
for cardiovascular diseases.
\\
\keywords{
        microRNA \and
        Machine Learning \and
        Myocardial Infarction \and
        Bioinformatics \and
        Biomarker \and
    }

@@

Cardiovascular diseases (CVDs) are the leading cause of human mortality,
accounting for 32\% of all global deaths. It is estimated that
approximately 85\% of CVD mortality is due to myocardial infarction (MI)
({``Cardiovascular Diseases ({CVDs})''} n.d.). MI is an acute coronary
syndrome characterized by sudden blockage and stenosis of the coronary
artery and subsequent myocardial ischemia, leading to extensive
cardiomyocyte damage and necrosis (Yap et al. 2023).

Over the last 50 years, numerous attempts have been made to use
biomarkers to facilitate diagnosis, assess the risk, follow-up therapy,
and determine therapeutic efficacy in CVD candidates. Based on released
guidelines, cardiac troponins (cTns) are used as a highly sensitive and
accurate approach for detecting MI. Despite these inherent advantages,
the high sensitivity of cTn-based assays has also led to more
false-positive results (Thygesen et al. 2018), necessitating the advent
and development of new modalities with pathological value. To improve
the diagnostic value of existing MI biomarkers, a combination of
complementary biological markers, such as microRNAs (miRNAs) and other
genetic factors, has been proposed. Previous research supports the
notion that miRNAs exhibit great potential as alternative biomarkers for
CVD detection and follow-up (Schulte et al. 2020). It has been suggested
that miRNAs possess 18-22 nucleotides and play a crucial role in the
regulation of gene expression. Evidence indicates that miRNAs are
involved in the pathogenesis of cardiac tissue injury (Schulte, Karakas,
and Zeller 2017). Several biological processes, such as angiogenesis,
cardiomyocyte growth and contractility, lipid metabolism, plaque
formation, and cardiac rhythm, are regulated by miRNAs (Kalayinia et al.
2021). These elements can easily circulate in biofluids and could be
considered theranostic targets in terms of CVDs (Schulte, Karakas, and
Zeller 2017). It has been postulated that the function and diagnostic
properties of miRNAs are beyond the myocardium in patients with CVD.
Specifically, the expression of miRNAs can vary in different biofluids
and cell components such as serum and peripheral blood mononuclear cells
(PBMCs) (Soler-Botija, Gálvez-Montón, and Bayés-Genís 2019).

PBMCs are a fraction of white blood cells, including monocytes,
lymphocytes, macrophages, and other cells of the immune system (Gao et
al. 2020). Emerging data indicate that PBMCs can be used as a valid
source of biomarkers for monitoring various pathological conditions. Of
note, the alteration of mRNAs and miRNAs under pathological conditions
provides valuable information about different kinds of disorders. PBMCs
can recapitulate the conditions of target tissues, thus providing a
highly sensitive and specific source of biomarkers (Mosallaei et al.
2022). Combined with these conditions, these cells are repositories of
dysregulated genes and miRNA expression profiles in CVDs (Gao et al.
2020; Mosallaei et al. 2022).

In recent years, the advent and application of machine learning (ML) has
been an exciting prospect for advancing scientific research. Although
the concept of ML and its initial algorithms were conceived many years
ago, recent improvements in computing power and access to vast amounts
of data have demonstrated that ML techniques outperform classical
statistical methods in various fields. Furthermore, the progress made in
omics technologies has enabled the analysis of massive and intricate
biological datasets, consisting of hundreds to thousands of samples,
which makes it possible for ML to extract valuable biological insights
and information from such data (Torun et al. 2023). Consequently, ML
provides innovative methods for merging and interpreting diverse types
of omics data, leading to the identification of new biomarkers. These
biomarkers can aid in precise disease prediction, patient
stratification, and the development of novel therapeutic approaches
(Reel et al. 2021).

In this study, we aimed to identify potential miRNA biomarkers in
patients with MI by combining and analyzing three different microarray
datasets from PBMCs. The integration of omics data with bioinformatics
and ML techniques could be a promising tool in the discovery of new and
more accurate biomarkers for monitoring MI. Additionally, this approach
can deepen our understanding of the underlying mechanisms of MI and aid
in the development of valid diagnostic biomarkers and patient
stratification.

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO)
database (\url{https://www.ncbi.nlm.nih.gov/geo/}). To obtain sufficient
classification power between MI, healthy, and CAD samples, a relatively
large sample size is needed Therefore, GSE59867 for the MI and CAD
samples and GSE56609 and GSE54475 for the healthy samples were selected.
All samples were produced using Affymetrix Human Gene 1.0 ST Array
platform (GPL6244). Only healthy, CAD, and early-stage MI samples were
selected from these datasets for further analysis. The basic information
for the three datasets evaluated in this study is provided in Table
@ref(tab:datasets). Bioinformatics analyses were conducted using R, ver.
4.2.0 (R Core Team 2022), RStudio (RStudio Team 2020). All plots and
graphics of these sections were created using the ggplot2 R package
(Wickham 2016).

\begin{longtable}[]{@{}cccccc@{}}
\caption{Sample information on the GEO microarray
datasets.}\tabularnewline
\toprule()
Dataset & Platform & Healthy & CAD & MI & Refrence \\
\midrule()
\endfirsthead
\toprule()
Dataset & Platform & Healthy & CAD & MI & Refrence \\
\midrule()
\endhead
GSE59867 & GPL6244 & - & 46 & 111 & (Maciejak et al. 2015) \\
GSE56609 & GPL6244 & 46 & - & - & (Matone et al. 2015) \\
GSE54475 & GPL6244 & 5 & - & - & (Canali et al. 2014) \\
\bottomrule()
\end{longtable}

\hypertarget{preprocessing}{%
\subsection{Preprocessing}\label{preprocessing}}

The raw data in the form of CEL files from all datasets were obtained
from GEO. To prepare the data for analysis, we utilized the fRMA package
(M. N. McCall, Bolstad, and Irizarry 2010) to facilitate preprocessing
of individual microarray samples and their consistent combination. For
each dataset, background correction was applied using the RMA algorithm,
followed by quantile normalization based on the reference distribution.
To account for probe-specific effects, batch effects were eliminated
during summarization and gene expression variances were estimated
accordingly. In cases where multiple probe sets matched the same gene,
the mean log-fold change was retained. Consequently, fRMA can serve as a
technique to remove batch effects across diverse datasets generated by
identical microarray platforms (Lazar et al. 2013). To ensure the
effectiveness of the batch effect removal, we employed principal
component analysis (PCA) and relative log expression (RLE) plots to
visualize the data before and after applying fRMA.

@@

The barcode algorithm was introduced by McCall et al. (Matthew N. McCall
et al. 2011), aimed to convert actual expression values into binary
barcode values. Extensive sample collections were gathered and
normalization was performed using fRMA across multiple platforms,
including the Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. By
utilizing these normalized datasets, the distribution of the observed
intensities for both the expressed and unexpressed genes was estimated.
The determination of whether a gene was expressed or not was based on
the following equation, where a value of 1 indicates expression and a
value of 0 indicates non-expression:

@@

In the barcode algorithm, the normalized intensity of gene \(i\) in
sample \(j\) is denoted as \(x_{ij}\). A user-defined parameter, \(C\),
was introduced along with the standard deviation (\(\sigma^{ne}\)) and
mean (\(\mu^{ne}\)) of the non-expressed distribution. Based on these
values, the barcode representation of a sample was generated as a vector
consisting of ones and zeros, representing the estimated expression
(ones) and non-expression (zeros) of each gene. The barcode function
within the R fRMA package was employed to implement the barcode
algorithm, utilizing the default value of \(C\).

To assess the differences in expressed ratios between the MI and healthy
control groups, Fisher's exact test was performed on the barcode values
of individual genes. Genes that exhibited a false discovery rate (FDR)
below \(0.05\), calculated using the Benjamini-Hochberg procedure to
account for multiple testing issues were identified as differentially
expressed genes (DEGs). The same procedures were applied to the CAD
versus healthy control comparison, as well as to the MI versus CAD
group, to identify DEGs specific to each comparison.

@@

The R clusterProfiler package (Yu et al. 2012) was utilized to perform
Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway enrichment
analysis and Gene Ontology (GO) functional annotation on the set of
DEGs. GO analysis encompassed three categories: biological process (BP),
cellular component (CC), and molecular function (MF). For statistical
significance, an adjusted p-value threshold of less than 0.05 was
employed. Enrichment analyses were conducted separately for DEGs
specific to the MI-healthy and CAD-healthy comparisons. All the default
parameters provided by the package were used in the analyses.

@@

ML analysis was performed using Python software, ver. 3.9, Numpy (Harris
et al. 2020), Pandas (McKinney 2010), and Scikit-Learn packages
(Pedregosa et al. 2011). Whenever hypertuning was needed, the Scikit-opt
package (Head et al. 2021) was used. In all ML analyses, the datasets
were divided into training and test sets at a 0.7:0.3 ratio, and all
reported results are the average of 10-fold cross-validation.

Two different approaches were used to select miRNAs for model training.
The first approach was to use differentially expressed miRNAs. In the
second approach, miRNAs with individual AUC-ROCs over 0.8 for separating
MI from CAD were selected. The results of these two approaches can
provide an informative comparison between the predictive capabilities of
sets of miRNAs selected with different logics.

@@

In this approach, a two-layer architecture is deployed to the data to
maximize the prediction values. The first layer predicted whether a
sample was healthy or not, and the second layer separated MI from CAD in
the samples that were predicted as not healthy in the first layer. To
this end, a distinct ML model was trained for each layer. Because, there
were a limited number of miRNAs in the DEGs, both layers were trained
with all of them. For further comparison with the models' performance,
the ROC curve of each miRNA for classifying healthy and not-healthy, as
well as CAD and MI, was generated using a logistic regression model.

\hypertarget{first-layer-for-the-isolation-of-healthy-and-not-healthy-samples}{%
\paragraph{First layer for the isolation of healthy and not-healthy
samples:}\label{first-layer-for-the-isolation-of-healthy-and-not-healthy-samples}}

A support vector machine (SVM) model using RBF kernels was trained and
hypertuned using all miRNAs in the DEGs. To handle the severe imbalance
in the number of samples (51 for the healthy group and 157 for the
not-healthy group), the sample weights for the healthy and the
not-healthy samples were set to 1 and 0.5, respectively. The ROC curve
and confusion matrix for the model are reported.

\hypertarget{second-layer-for-separating-the-mi-and-cad-samples}{%
\paragraph{Second layer for separating the MI and CAD
samples:}\label{second-layer-for-separating-the-mi-and-cad-samples}}

Different models were investigated to achieve the highest classification
performance. To do so, SVM (with linear, polynomial, and RBF kernels),
logistic regression (LR), random forests (RF), k-nearest neighbor (kNN),
gradient boosting (GB), XGBoost (XGB) and decision tree (DT) models were
trained. All models were trained with their preset parameters using
10-fold cross-validation. The criteria for selecting the best model were
the highest accuracy and AUC-ROC for the test set. The best model was
hypertuned using the scikit-opt package (Head et al. 2021) for the best
classification performance. The ROC curve and confusion matrix for the
best model are reported.

@@

As in the previous approach, a two-layer strategy was employed. The
first layer classified samples into healthy and not-healthy, and the
second layer separated the MI and CAD samples. However, to keep the
number of miRNAs as low as possible, miRNAs were selected from the
second layer and their performance was evaluated in the first layer. The
AUC-ROC of all miRNAs for classifying MI and CAD samples was calculated,
and miRNAs with AUC-ROC \(> 0.8\) were selected. ROC curves for each
selected miRNA for separating healthy samples from not-healthy samples
and MI from CAD samples were also plotted for further comparison.

\hypertarget{first-layer-for-the-isolation-of-healthy-and-not-healthy-samples-1}{%
\paragraph{First layer for the isolation of healthy and not-healthy
samples:}\label{first-layer-for-the-isolation-of-healthy-and-not-healthy-samples-1}}

An SVM model with an RBF kernel is trained using the selected set of
miRNAs. Additionally, the model was hypertuned to find the
hyperparameters for the highest AUC-ROC and accuracy. The same sample
weights as in the previous approach (1 for healthy and 0.5 for
not-healthy samples) were used. The ROC curve and confusion matrix for
the model were reported.

\hypertarget{second-layer-for-separating-the-mi-and-cad-samples-1}{%
\paragraph{Second layer for separating the MI and CAD
samples:}\label{second-layer-for-separating-the-mi-and-cad-samples-1}}

The selected miRNA set was used to train different algorithms to
determine the best model. Similar to the previous approach, the SVM
(with linear, polynomial, and RBF kernels), LR, RF, kNN, GB, XGB, and DT
models were trained. All models were trained with their preset
parameters using 10-fold cross-validation. The models with the highest
AUC-ROC and accuracy on the test set were selected and hypertuned using
the scikit-opt package (Head et al. 2021). The ROC curve and confusion
matrix for the best model were reported.

@@

\hypertarget{preprocessing-1}{%
\subsection{Preprocessing}\label{preprocessing-1}}

The PCA plots of the samples are shown in Figures @ref(fig:PCA)A and B.
Healthy samples were separated from the CAD or MI samples in the primary
data and after conducting fRMA. In the RLE plot, there was a distinct
difference between the dataset means for all samples before fRMA was
performed (Figure @ref(fig:PCA)C). All datasets were rearranged to
approximately 0 in the RLE plot after fRMA was conducted (Figure
@ref(fig:PCA)D). Moreover, there was an apparent change in the
interquantile distances, but the values were still greater than 0.1.

\begin{figure}
\includegraphics[width=1\linewidth]{../../DEG and DEM/Pics/PCA and RLE of trains} \caption{Principal component analysis plots for (A) primary data and (B) the data after fRMA, and the relative log expression plots for (C) primary data and (D) the data after fRMA.}\label{fig:PCA}
\end{figure}

@@

According to the cutoff criterion of \(FDR < 0.05\), there were 860 DEGs
between MI and healthy samples. Among them, 323 were up-regulated, and
537 were down-regulated in the MI group compared to the healthy group.
In the CAD and healthy group comparison, we found 670 DEGs, of which 262
and 408 DEGs were up- and down-regulated, respectively, in CAD samples.
In the MI and CAD groups, the number of DEGs was 260, and the numbers of
up- and down-regulated genes in MI samples were 144 and 116,
respectively, compared to CAD samples. The data are summarized in Table
@ref(tab:DEGstab).

@@

{\centering \includegraphics[width=0.45\linewidth]{Complete-V1_files/figure-latex/venn-1}

@@

\caption{Venn diagram for DEGs in CAD/Healthy, MI/Healthy, and MI/CAD.}\label{fig:venn}
\end{figure}

The Venn diagram in Figure @ref(fig:venn) shows that the CAD and MI
samples shared most of their DEGs. From 860 DEGs of MI/healthy and 670
DEGs of CAD/healthy, 531 genes were common, which is 62\% of MI/healthy
DEGs and 79\% of CAD/healthy DEGs.

\hypertarget{go-and-kegg-enrichment-analyses-of-the-degs}{%
\subsection{GO and KEGG enrichment analyses of the
DEGs}\label{go-and-kegg-enrichment-analyses-of-the-degs}}
